---
title: "Soil_carbon_predictiion"
output: 
date: "2025-04-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Load libraries
library(dplyr)
library(lubridate)
library(ggplot2)
library(randomForest) 
library(caTools)
library(caret)
library(ggpmisc)
```
The data used here includes:-

The daily soil CO2 flux data observed from LICOR's Automated Soil COâ‚‚ Flux System

The variables soil water content (SWC), soil temperature (T_soil), atmospheric temperature (T_air), relative humidity (RH), incoming long-wave (RAD_lw) and short-wave radiation (RAD_sw), and soil heat flux (SHF) and the difference between atmospheric and soil temperatures (T_diff)

Hourly data from 2015 to 2016 were used for training and testing the models. Variables from 2018 for prediction

```{r}
#Data

df_predict<- read.csv("D:/dattebayo/flux/soil_github/soil_github/test_data.csv")
df_train<- read.csv("D:/dattebayo/flux/soil_github/soil_github/training.csv")


# Ensure dates are in proper format
df_train$date<-as.POSIXct(df_train$date, "%d-%m-%Y %H:%M", tz="")
head(df_train)
```
```{r}
# Ensure dates are in proper format
df_predict$date<-as.POSIXct(df_predict$date, "%d-%m-%Y %H:%M", tz="")
head(df_predict)
```

```{r}
# Splitting data for training and testing of models
set.seed(25) # for reproducibility
split <- sample.split(df_train, SplitRatio = 0.7) 
train <- subset(df_train, split == "TRUE") 
test <- subset(df_train, split == "FALSE") 
```


## Q10 Model

Temperature sensitivity, Q10, model is commonly used to estimate soil carbon flux based on the observed temperature dependency on biological activity.

Rs= ð›¼exp(Î² T_soil)

where Rs is the soil respiration

```{r}

# Calculate logarithmic values of soil respiration
train$log_respiration <- log(train$co2_flux)
train<-filter(train, !is.na(train$log_respiration)) # removing NAs


# Wet Data
wet_data<- train %>% filter(SWC > 0.5)

# Fitting the model
fit_wet <- lm(log_respiration ~ T_soil, data = wet_data)

# Model summary
summary(fit_wet)

# Extract the slope of the regression line
al_wet <- coef(fit_wet)[1]
alpha_wet<- exp(al_wet)
beta_wet <-coef(fit_wet)[2]

cat("Parameter alpha for wet state is ", alpha_wet, "\n")
cat("Parameter beta for wet state is ", beta_wet, "\n")

```
```{r}
# Dry data
dry_data<- train %>% filter(SWC <= 0.5)

#Fitting the model
fit_dry <- lm(log_respiration ~ T_soil, data = dry_data)

# Model summary
summary(fit_dry)

# Extract the slope of the regression line
al_dry <- coef(fit_dry)[1]
alpha_dry<- exp(al_dry)
beta_dry <-coef(fit_dry)[2]

cat("Parameter alpha for dry state is ", alpha_dry, "\n")
cat("Parameter beta for dry state is ", beta_dry, "\n")

```


Parameters of Q10 model calculated for Betul region
ð›¼wet 0.196
Î²wet 0.080
ð›¼dry 2.156
Î²dry -0.003


```{r}

# Calculate soil respiration using Q10 model on test sample
q10_model<- test %>% mutate(respiration = case_when(SWC > 0.5 ~ alpha_wet * exp(beta_wet * T_soil),
                                                        SWC <= 0.5 ~ alpha_dry * exp(beta_dry * T_soil),
                                                        TRUE ~ NA))

postResample(pred = q10_model$respiration, obs = q10_model$co2_flux) 
```
```{r}
# Plot regression line
ggplot(q10_model,aes(x=co2_flux,y=respiration))+
  geom_point(color="#19285f")+ 
  theme_bw()+xlab("Observed value")+ylab("Predicted value")+  
  stat_poly_line(formula = y ~ x, method = "lm", se = FALSE, color = "black")+
  stat_poly_eq()+
  theme(axis.text=element_text(size=20), axis.title=element_text(size=22,face="bold"))


```

```{r}

# Predict soil respiration for 2018 
q10<- df_predict %>% mutate(respiration = case_when(SWC > 0.5 ~ alpha_wet * exp(beta_wet * T_soil),
                                                        SWC <= 0.5 ~ alpha_dry * exp(beta_dry * T_soil),
                                                        TRUE ~ NA)) %>%
  group_by(Date= date(date)) %>%   
  summarise(respiration=mean(respiration)*12.01*60*60*24/1000000 )


# Plot the predicted values

ggplot(q10, aes(x = as.Date(Date), group = cumsum(c(0, diff(Date) > 1)))) +
  geom_path(aes(y = respiration)) +
  scale_x_date(name = "Date", breaks = "1 month", date_labels = "%b") +
  ylab(expression("Soil Carbon Flux (gC m"^-2*".day"^-1*")")) +
  theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 18, face = "bold")) +
  ylim(c(0,5))
```

## MODIFIED Q10

Takes into consideration soil moisture as well soil temperature

 Rs= Î²0.e(Î²1.T_soil).e(Î²2.SWC +Î²3.SWC^2)

```{r}
# Model
mq10_model <- lm(log_respiration ~ T_soil + SWC + I(SWC^2), data = train)

summary(mq10_model)
# Extract coefficients
beta_0_prime <- coef(mq10_model)[1]  # ln(beta_0)
beta_1 <- coef(mq10_model)[2]    # beta_1
beta_2 <- coef(mq10_model)[3]    # beta_2
beta_3 <- coef(mq10_model)[4]    # beta_3
beta_0 <- exp(beta_0_prime)      # beta_0

cat("Parameter beta 0 is ", beta_0, "\n")
cat("Parameter beta 1 is ", beta_1, "\n")
cat("Parameter beta 2 is ", beta_2, "\n")
cat("Parameter beta 3 is ", beta_3, "\n")

```



The parameters for modified Q10 model
Î²0 0.00025
Î²1 -0.005
Î²2 33.92
Î²3 -29.887


```{r}
# Test modified Q10
modified_q10<-test %>% mutate(respiration = beta_0 * exp(beta_1 * T_soil) * exp(beta_2 * SWC) * exp(beta_3 * (SWC^2)))

postResample(pred = modified_q10$respiration, obs = modified_q10$co2_flux) 

```

```{r}
# Scatterplot with regression line
ggplot(modified_q10,aes(x=co2_flux,y=respiration))+
  geom_point(color="#19285f")+ 
  theme_bw()+xlab("Observed value")+ylab("Predicted value")+  
  stat_poly_line(formula = y ~ x, method = "lm", se = FALSE, color = "black")+
  stat_poly_eq()+
  theme(axis.text=element_text(size=20), axis.title=element_text(size=22,face="bold"))
```

```{r}

# Predict soil repiration using modified Q10
mod_q10<- df_predict %>% mutate(respiration = beta_0 * exp(beta_1 * T_soil) * exp(beta_2 * SWC) * exp(beta_3 * (SWC^2))) %>% 
  group_by(Date= date(date)) %>%   summarise(respiration= mean(respiration)*12.01*60*60*24/1000000 )

# Plotting data
ggplot(mod_q10, aes(x = as.Date(Date), group = cumsum(c(0, diff(Date) > 1)))) +
  geom_path(aes(y = respiration)) +
  scale_x_date(name = "Date", breaks = "1 month", date_labels = "%b") +
  ylab(expression("Soil Carbon Flux (gC m"^-2*".day"^-1*")")) +
  theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 18, face = "bold"))+
  ylim(c(0,5))
```


## RF2

Random forest model using T_soil and SWC as variables

```{r}
set.seed(394)
# Ten fold cross validation
train_control<- trainControl(method="cv", number=10, savePredictions = TRUE)
set.seed(24) # for reproducibility

# Training data
rf2_model <-caret::train(x = train[,c(3,4)], 
                        y = train[,2], trControl=train_control, method="rf")

print(rf2_model)

```
```{r}
# Test the data
rf2_p <- predict(rf2_model, newdata = test[,c(3,4)]) 
rf2_o<- test[,2]

postResample(pred = rf2_p, obs = rf2_o) 

```
```{r}
# Plot regression line
ggplot(data.frame(obs=rf2_o, pred=rf2_p),aes(x=obs,y=pred))+
  geom_point(color="#19285f")+ 
  theme_bw()+xlab("Observed value")+ylab("Predicted value")+  
  stat_poly_line(formula = y ~ x, method = "lm", se = FALSE, color = "black")+
  stat_poly_eq()+
  theme(axis.text=element_text(size=20), axis.title=element_text(size=22,face="bold"))
```

```{r}
# Prediction 

rf2_pred <- predict(rf2_model, newdata = df_predict[,c(2,3)]) 
rf2_predicted<- cbind(df_predict, rf2_pred)

rf2_predict <- rf2_predicted %>% group_by(Date= date(date)) %>%   summarise( rf2_pred= mean(rf2_pred)*12.01*60*60*24/1000000 )

# Plot predicted values

ggplot(rf2_predict, aes(x = as.Date(Date), group = cumsum(c(0, diff(Date) > 1)))) +
  geom_path(aes(y = rf2_pred)) +
  scale_x_date(name = "Date", breaks = "1 month", date_labels = "%b") +
  ylab(expression("Soil Carbon Flux (gC m"^-2*".day"^-1*")")) +
  theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 18, face = "bold"))+
  ylim(c(0,5))

```


## RF8

Random forest model using all the variables mentioned in the data

```{r}
# Training data

set.seed(245) # for reproducibility
rf8_model <-caret::train(x = train[,-c(1,2,11)], 
                        y = train[,2], trControl=train_control, method="rf")

print(rf8_model)
```

```{r}
# Test the data
rf8_p <- predict(rf8_model, newdata = test[,-c(1,2,11)]) 
rf8_o<- test[,2]

postResample(pred = rf8_p, obs = rf8_o) 
```

```{r}
# Plot regression line
ggplot(data.frame(obs=rf8_o, pred=rf8_p),aes(x=obs,y=pred))+
  geom_point(color="#19285f")+ 
  theme_bw()+xlab("Observed value")+ylab("Predicted value")+  
  stat_poly_line(formula = y ~ x, method = "lm", se = FALSE, color = "black")+
  stat_poly_eq()+
  theme(axis.text=element_text(size=20), axis.title=element_text(size=22,face="bold"))
```

```{r}
# Prediction for 2018
rf8_pred <- predict(rf8_model, newdata = df_predict[,-c(1)]) 
rf8_predicted<- cbind(df_predict, rf8_pred)

rf8_predict <- rf8_predicted %>% group_by(Date= date(date)) %>% summarise( rf8_pred= mean(rf8_pred)*12.01*60*60*24/1000000 )

#Plot predicted data

ggplot(rf8_predict, aes(x = as.Date(Date), group = cumsum(c(0, diff(Date) > 1)))) +
  geom_path(aes(y = rf8_pred)) +
  scale_x_date(name = "Date", breaks = "1 month", date_labels = "%b") +
  ylab(expression("Soil Carbon Flux (gC m"^-2*".day"^-1*")")) +
  theme_bw() +
  theme(axis.text = element_text(size = 16),
        axis.title = element_text(size = 18, face = "bold")) +
  ylim(0,5)

```


```{r}
plot(varImp(rf8_model, main = "Variable Importance - Random Forest"))
```
